---
title: "Bhuyan_Chuang_Martinez"
author: "Jessalyn Chuang, Mazhar Bhuyan, Sayra Martinez"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

Github repository: https://github.com/jessalynlc/BhuyanChuangMartinez_ENV797_TSA_ForecastCompetition_S25

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4
)

```

#Data Wrangling
```{r package, message=FALSE, warning=FALSE, include=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(zoo)
library(kableExtra)
library(readxl)
library(xts)
```

#Data Import and Primary Cleaning
```{r Data Wrangling, include=FALSE}
#Importing time series data (relative humidity and temperature optional but importing)
#load data: this is daily data, with 24 different sensors
# Load data
load_data <- read_excel(here::here("Data", "load.xlsx"))
relative_humidity_data <- read_excel(here::here("Data", "relative_humidity.xlsx"))
temperature_data <- read_excel(here::here("Data", "temperature.xlsx"))

# load_data <- read_excel(
#   path = "./Data/load.xlsx",
#   sheet = 1
# )
sum(is.na(load_data))
sum(load_data == 0, na.rm = TRUE)

# relative_humidity_data <- read_excel(
#   path = "./Data/relative_humidity.xlsx",
#   sheet = 1
# )
sum(is.na(relative_humidity_data))
sum(relative_humidity_data == 0, na.rm = TRUE)

# temperature_data <- read_excel(
#   path = "./Data/temperature.xlsx",
#   sheet = 1
# )
sum(is.na(temperature_data))
sum(temperature_data == 0, na.rm = TRUE)
```

```{r include=FALSE}
load_processed <- load_data %>%
  pivot_longer(cols = starts_with("h"), names_to = "hour", values_to = "electricity_demand") %>%
  mutate(hour = as.integer(sub("h", "", hour)), date = ymd(date), year = year(date), 
         month = month(date), day = day(date)) %>% 
  select(date, year, month, day, hour, electricity_demand)

humidity_processed <- relative_humidity_data %>%
  pivot_longer(cols = starts_with("rh"), names_to = "hour", values_to = "relative_humidity") %>%
  mutate(hour = as.integer(gsub("[^0-9]", "", hour)), date = ymd(date), year = year(date), 
         month = month(date), day = day(date)) %>% 
  select(date, year, month, day, hour, relative_humidity)

temp_processed <- temperature_data %>%
  pivot_longer(cols = starts_with("t"), names_to = "hour", values_to = "temperature") %>%
  mutate(hour = as.integer(gsub("[^0-9]", "", hour)), date = ymd(date), year = year(date), 
         month = month(date), day = day(date)) %>% 
  select(date, year, month, day, hour, temperature)
```

## Converting to daily by taking averages
```{r include=FALSE}
daily_load <- load_processed %>%
  group_by(date) %>%
  summarise(daily_avg_load = mean(electricity_demand, na.rm = TRUE)) %>%
  ungroup()

daily_humidity <- humidity_processed %>%
  group_by(date) %>%
  summarise(daily_avg_humidity = mean(relative_humidity, na.rm = TRUE)) %>%
  ungroup()

daily_temp <- temp_processed %>%
  group_by(date) %>%
  summarise(daily_avg_temp = mean(temperature, na.rm = TRUE)) %>%
  ungroup()
```

## Merging the daily datasets to a full daily dataset 
```{r include=FALSE}
full_daily <- daily_load %>%
  inner_join(daily_temp, by = "date") %>%
  inner_join(daily_humidity, by = "date") %>%
  arrange(date)
head(full_daily)
```

##Converting to time series object
```{r message=FALSE, warning=FALSE}
ts_electricity_daily <- msts(full_daily$daily_avg_load, 
                              seasonal.periods = c(7, 365.25), 
                              start = decimal_date(as.Date("2005-01-01")))

autoplot(ts_electricity_daily) + ggtitle("Electricity Demand: Daily")

ts_daily_train <- window(ts_electricity_daily, end = c(2009, 365))
ts_daily_test  <- window(ts_electricity_daily, start = c(2010, 1), end = c(2010, 59))

autoplot(ts_daily_train) + ggtitle("Training Set: Daily Demand (2005-2009)")
autoplot(ts_daily_test) + ggtitle("Test Set: Daily Demand (Jan-Feb 2010)")
```


### Model 1: NNAR + Fourier (K = c(2,8)) on Train/Test Data
```{r nnar_fourier_k28, message=FALSE, warning=FALSE}
horizon <- length(ts_daily_test)

NN_fit_k28 <- nnetar(ts_daily_train,
                  p = 2,
                  P = 2,
                  xreg = fourier(ts_daily_train, K = c(2, 8)))

NN_for_k28 <- forecast(NN_fit_k28, h = horizon,
                       xreg = fourier(ts_daily_train, K = c(2, 8), h = horizon))

autoplot(ts_daily_train) +
   autolayer(NN_for_k28, series = "NNAR K=2,8", PI = FALSE) +
   ylab("Load")

accuracy(NN_for_k28, ts_daily_test)
```

### Model 2: NNAR + Fourier (K = c(2,12)) on Train/Test Data (Baseline)
```{r nnar_fourier_k212_baseline, message=FALSE, warning=FALSE}
NN_fit_k212_base <- nnetar(ts_daily_train,
                  p = 2,
                  P = 2,
                  xreg = fourier(ts_daily_train, K = c(2, 12)))

NN_for_k212_base <- forecast(NN_fit_k212_base, h = horizon,
                             xreg = fourier(ts_daily_train, K = c(2, 12), h = horizon))

autoplot(ts_daily_train) +
   autolayer(NN_for_k212_base, series = "NNAR K=2,12", PI = FALSE) +
   ylab("Load")

accuracy(NN_for_k212_base, ts_daily_test)
```

### Model 3: NNAR + Fourier on Train/Test Data
```{r nnar_fourier_k212, message=FALSE, warning=FALSE}

# NNAR + Fourier Model
# I chose to use a Neural Network Autoregressive (NNAR) model with Fourier terms (K = c(2, 12))
# to capture complex seasonal and nonlinear patterns in the daily electricity data.
# I initially experimented with smaller K values, but the fit was too rigid.
# Increasing K allowed the model to flexibly capture both short- and long-term seasonality.

horizon <- length(ts_daily_test)

# Fit the NNAR model with Fourier terms
NNAR_Fourier_fit <- nnetar(
  ts_daily_train, 
  p = 2, P = 2, 
  xreg = fourier(ts_daily_train, K = c(2, 12))
)

# Forecast using the fitted model
NNAR_Fourier_forecast <- forecast(
  NNAR_Fourier_fit, 
  h = horizon, 
  xreg = fourier(ts_daily_train, K = c(2, 12), h = horizon)
)

# Plot the forecast against the test data
# This visual check helped me confirm that the forecast captured both the trend and seasonal fluctuation well.
autoplot(ts_daily_test) + 
  autolayer(NNAR_Fourier_forecast, series = "NNAR + Fourier (K=2,12)")

# Calculate accuracy metrics

accuracy(NNAR_Fourier_forecast, ts_daily_test) # Score 23.48



```

### Model 4: NNAR + Fourier (K = c(3,18)) on Train/Test Data
```{r nnar_fourier_k318_test, message=FALSE, warning=FALSE}
horizon <- length(ts_daily_test)
K <- c(3, 18)

xreg_train <- fourier(ts_daily_train, K = K)
xreg_test  <- fourier(ts_daily_train, K = K, h = horizon)

NN_fit_k318 <- nnetar(
  ts_daily_train,
  p    = 2,
  P    = 2,
  xreg = xreg_train,
  size = 10,
  decay = 0.01,
  maxNWts = 2000
)

NN_for_k318 <- forecast(
  NN_fit_k318,
  h    = horizon,
  xreg = xreg_test
)

autoplot(NN_for_k318) + ggtitle("Model 4: NNAR + Fourier (K = c(3,18)) Forecast on Test Data")
accuracy(NN_for_k318, ts_daily_test) # Score 22.68
```


### Model 5: TBATS Model
```{r tbats_model, message=FALSE, warning=FALSE}
# Fit TBATS model (training data assumed to be ts_daily_train)
TBATS_fit <- tbats(ts_daily_train)

TBATS_for <- forecast(TBATS_fit, h = horizon)

# Forecast plot
autoplot(TBATS_for) +
  ggtitle("Model 5: TBATS Forecast") +
  ylab("Load")

# Overlay plot
autoplot(ts_daily_train) +
  autolayer(TBATS_for, series = "TBATS", PI = FALSE) +
  ylab("Load")

# Accuracy on test set
accuracy(TBATS_for, ts_daily_test)
```



### Model Comparison: NNAR + Fourier + TBATS

```{r compare-nnar-models, message=FALSE, warning=FALSE}


# Compare performance across all six models

comparison <- data.frame(
  Model = c(
    "Model 1: NNAR + Fourier (K = c(2,8))",
    "Model 2: NNAR + Fourier (K = c(2,12)) Baseline",
    "Model 3: NNAR + Fourier (K = c(2,12)) on Test",
    "Model 4: NNAR + Fourier (K = c(3,18))",
    "Model 5: TBATS"
  ),
  RMSE = c(
    accuracy(NN_for_k28, ts_daily_test)[2, "RMSE"],
    accuracy(NN_for_k212_base, ts_daily_test)[2, "RMSE"],
    accuracy(NNAR_Fourier_forecast, ts_daily_test)[2, "RMSE"],
    accuracy(NN_for_k318, ts_daily_test)[2, "RMSE"],
    accuracy(TBATS_for, ts_daily_test)[2, "RMSE"]
  ),
  MAE = c(
    accuracy(NN_for_k28, ts_daily_test)[2, "MAE"],
    accuracy(NN_for_k212_base, ts_daily_test)[2, "MAE"],
    accuracy(NNAR_Fourier_forecast, ts_daily_test)[2, "MAE"],
    accuracy(NN_for_k318, ts_daily_test)[2, "MAE"],
    accuracy(TBATS_for, ts_daily_test)[2, "MAE"]
  ),
  MAPE = c(
    accuracy(NN_for_k28, ts_daily_test)[2, "MAPE"],
    accuracy(NN_for_k212_base, ts_daily_test)[2, "MAPE"],
    accuracy(NNAR_Fourier_forecast, ts_daily_test)[2, "MAPE"],
    accuracy(NN_for_k318, ts_daily_test)[2, "MAPE"],
    accuracy(TBATS_for, ts_daily_test)[2, "MAPE"]
  )
)

# Nicely formatted table
kable(
  comparison, 
  caption = "Performance Comparison of All 5 Models (Train/Test Evaluation)",
  digits = 3
)

```

# Retraining and Forecasting on Full Data


### Model 1: NNAR + Fourier (K = c(2,8))

```{r}
# Use horizon for forecasting
horizon <- length(ts_daily_test)

# Fit the NNAR model with Fourier terms using the full training dataset
NN_fit_k28_full <- nnetar(ts_electricity_daily, p = 2, P = 2, xreg = fourier(ts_electricity_daily, K = c(2, 8)))

# Forecast for the desired horizon
NN_for_k28_full <- forecast(NN_fit_k28_full, h = horizon, xreg = fourier(ts_electricity_daily, K = c(2, 8), h = horizon))

# Create a data frame for the forecast results
forecast_results_k28 <- data.frame(
  date = seq(as.Date("2011-01-01"), by = "day", length.out = horizon),
  load = NN_for_k28_full$mean
)

# View the forecast results
head(forecast_results_k28)

```


### Model 2: NNAR + Fourier (K = c(2,12))

```{r}
# Use horizon for forecasting
horizon <- length(ts_daily_test)

# Fit the NNAR model with Fourier terms using the full training dataset
NN_fit_k212_base_full <- nnetar(ts_electricity_daily, p = 2, P = 2, xreg = fourier(ts_electricity_daily, K = c(2, 12)))

# Forecast for the desired horizon
NN_for_k212_base_full <- forecast(NN_fit_k212_base_full, h = horizon, xreg = fourier(ts_electricity_daily, K = c(2, 12), h = horizon))

# Create a data frame for the forecast results
forecast_results_k212_base <- data.frame(
  date = seq(as.Date("2011-01-01"), by = "day", length.out = horizon),
  load = NN_for_k212_base_full$mean
)

# View the forecast results
head(forecast_results_k212_base)

```



### Model 3: NNAR + Fourier

```{r}
# Fit the NNAR model with Fourier terms using the full training dataset
NNAR_Fourier_full_fit <- nnetar(ts_electricity_daily, p = 2, P = 2, xreg = fourier(ts_electricity_daily, K = c(2, 12)))

# Forecast for the desired horizon (January 1 to February 28, 2011)
horizon_full <- 59  # Number of days to forecast
NNAR_Fourier_full_forecast <- forecast(NNAR_Fourier_full_fit, h = horizon_full, xreg = fourier(ts_electricity_daily, K = c(2, 12), h = horizon_full))

# Create a data frame for the forecast results
forecast_results <- data.frame(
  date = seq(as.Date("2011-01-01"), by = "day", length.out = horizon_full),
  load = NNAR_Fourier_full_forecast$mean
)

```

### Model 4: NNAR + Fourier (K = c(3,18))
```{r}
# Set K for Fourier
K4 <- c(3, 18)
# Training
ts_full <- window(ts_electricity_daily, end = c(2010, 365))

# Create Fourier regressors
xreg_full_4 <- fourier(ts_full, K = K4)
xreg_fc_4   <- fourier(ts_full, K = K4, h = horizon)

# Fit the NNAR model with extended size and decay
fit_nnar_k318 <- nnetar(
  ts_full,
  p = 2,
  P = 2,
  xreg = xreg_full_4,
  size = 10,
  decay = 0.01,
  maxNWts = 2000
)

# Forecast
fc_nnar_k318 <- forecast(
  fit_nnar_k318,
  h = horizon,
  xreg = xreg_fc_4
)

# Create forecast dataframe
final_nnar_k318_df <- data.frame(
  date = seq(as.Date("2011-01-01"), by = "day", length.out = horizon_full),
  load = as.numeric(fc_nnar_k318$mean)
)

```

### Model 5: TBATS
```{r}
# Use horizon for forecasting
horizon <- length(ts_daily_test)

# Fit TBATS model (training data assumed to be ts_daily_train)
TBATS_fit_full <- tbats(ts_daily_train)

# Forecast for the desired horizon
TBATS_for_full <- forecast(TBATS_fit_full, h = horizon)

# Forecast plot
autoplot(TBATS_for_full) +
  ggtitle("Model 5: TBATS Forecast") +
  ylab("Load")

# Overlay plot
autoplot(ts_daily_train) +
  autolayer(TBATS_for_full, series = "TBATS", PI = FALSE) +
  ylab("Load")

# Accuracy on test set
accuracy(TBATS_for_full, ts_daily_test)

# Create a data frame for the forecast results
forecast_results_tbats <- data.frame(
  date = seq(as.Date("2011-01-01"), by = "day", length.out = horizon),
  load = as.numeric(TBATS_for_full$mean)
)

```


# Exporting Dataset

```{r}
# write.csv(forecast_results_k28, file = here::here("Submission Files", "NNAR_Fourier_K28.csv"), row.names = FALSE)
# write.csv(forecast_results_k212_base, file = here::here("Submission Files", "forecast_results_k212_base.csv"), row.names = FALSE)
# write.csv(forecast_results, file = here::here("Submission Files", "NNAR_Fourier.csv"), row.names = FALSE)
# write.csv(final_nnar_k318_df,file = here::here("Submission Files", "submission_nnar_k318.csv"), row.names = FALSE)
# write.csv(forecast_results_tbats,file = here::here("Submission Files", "forecast_results_tbats.csv"), row.names = FALSE)
# write.csv(forecast_results_tbats, "submission_arima_nnar_hybrid.csv", row.names = FALSE)

```















<!-- ## Model Testing Not Included -->
<!-- ### 1. NNAR + Fourier -->

<!-- ```{r} -->
<!-- # NNAR + Fourier Model -->
<!-- # I chose to use a Neural Network Autoregressive (NNAR) model with Fourier terms (K = c(2, 12)) -->
<!-- # to capture complex seasonal and nonlinear patterns in the daily electricity data. -->
<!-- # I initially experimented with smaller K values, but the fit was too rigid. -->
<!-- # Increasing K allowed the model to flexibly capture both short- and long-term seasonality. -->

<!-- horizon <- length(ts_daily_test) -->

<!-- # Fit the NNAR model with Fourier terms -->
<!-- NNAR_Fourier_fit <- nnetar( -->
<!--   ts_daily_train, -->
<!--   p = 2, P = 2, -->
<!--   xreg = fourier(ts_daily_train, K = c(2, 12)) -->
<!-- ) -->

<!-- # Forecast using the fitted model -->
<!-- NNAR_Fourier_forecast <- forecast( -->
<!--   NNAR_Fourier_fit, -->
<!--   h = horizon, -->
<!--   xreg = fourier(ts_daily_train, K = c(2, 12), h = horizon) -->
<!-- ) -->

<!-- # Plot the forecast against the test data -->
<!-- # This visual check helped me confirm that the forecast captured both the trend and seasonal fluctuation well. -->
<!-- autoplot(ts_daily_test) + -->
<!--   autolayer(NNAR_Fourier_forecast, series = "NNAR + Fourier (K=2,12)") -->

<!-- # Calculate accuracy metrics -->

<!-- accuracy(NNAR_Fourier_forecast, ts_daily_test) # Score 23.48 -->
<!-- ``` -->

<!-- ```{r} -->


<!-- # Export the forecast results to a CSV file -->


<!-- # Display the forecast results -->
<!-- head(forecast_results) -->
<!-- ``` -->

<!-- ## Not so Great -->
<!-- ### NN + Fourier K(2,12) with regressor -->
<!-- ```{r} -->
<!-- ts_temp_daily <- msts(full_daily$daily_avg_temp, -->
<!--                       seasonal.periods = c(7, 365.25), -->
<!--                       start = decimal_date(as.Date("2005-01-01"))) -->

<!-- # Splitting temperature into train/test sets to match electricity demand -->
<!-- temp_train <- window(ts_temp_daily, end = c(2009, 365)) -->
<!-- temp_test  <- window(ts_temp_daily, start = c(2010, 1), end = c(2010, 59)) -->


<!-- horizon <- length(ts_daily_test) -->
<!-- K <- c(2, 12) -->

<!-- xreg_train <- cbind( -->
<!--   temp_train, -->
<!--   fourier(ts_daily_train, K = K) -->
<!-- ) -->
<!-- xreg_test <- cbind( -->
<!--   temp_test, -->
<!--   fourier(ts_daily_train, K = K, h = horizon) -->
<!-- ) -->

<!-- NN_fit_k212 <- nnetar( -->
<!--   ts_daily_train, -->
<!--   p    = 2, -->
<!--   P    = 2, -->
<!--   xreg = xreg_train -->
<!-- ) -->

<!-- NN_for_k212 <- forecast( -->
<!--   NN_fit_k212, -->
<!--   h    = horizon, -->
<!--   xreg = xreg_test -->
<!-- ) -->

<!-- autoplot(NN_for_k212) + ggtitle("NNAR + Fourier (K = c(2,12)) Forecast") -->
<!-- accuracy(NN_for_k212, ts_daily_test) ## Score 42 -->
<!-- ``` -->


<!-- ### Model 6: ARIMA + NNAR Hybrid on Train/Test Data -->

<!-- ```{r} -->
<!-- # Convert to univariate ts with daily frequency (e.g., yearly seasonality only) -->
<!-- # 1. Make sure your input is a plain ts object -->
<!-- ts_daily_train_ts <- ts(as.numeric(ts_daily_train), start = c(2005, 1), frequency = 365.25) -->

<!-- # 2. Fit plain ARIMA model -->
<!-- fit_arima <- auto.arima(ts_daily_train_ts, seasonal = TRUE) -->

<!-- # 3. Forecast ARIMA -->
<!-- fc_arima_test <- forecast(fit_arima, h = horizon) -->

<!-- # 4. Get ARIMA residuals -->
<!-- res_arima <- residuals(fit_arima) -->

<!-- # 5. Fit NNAR on residuals -->
<!-- fit_nnar_resid <- nnetar(res_arima) -->

<!-- # 6. Forecast residuals -->
<!-- fc_nnar_resid_test <- forecast(fit_nnar_resid, h = horizon) -->

<!-- # 7. Combine ARIMA + NNAR forecasts -->
<!-- fc_hybrid_test <- fc_arima_test$mean + fc_nnar_resid_test$mean -->


<!-- autoplot(ts_daily_test) + -->
<!--   autolayer(fc_arima_test$mean, series = "ARIMA", PI = FALSE) + -->
<!--   autolayer(fc_hybrid_test, series = "ARIMA + NNAR", PI = FALSE) + -->
<!--   ggtitle("Model 6: ARIMA + NNAR Hybrid Forecast") + -->
<!--   ylab("Load") -->


<!-- fc_hybrid_test <- fc_arima_test$mean + fc_nnar_resid_test$mean -->

<!-- # Convert hybrid forecast to a time series to match test set -->
<!-- fc_hybrid_ts <- ts(fc_hybrid_test, start = start(ts_daily_test), frequency = frequency(ts_daily_test)) -->

<!-- # Compute accuracy -->
<!-- accuracy_hybrid <- accuracy(fc_hybrid_ts, ts_daily_test) -->
<!-- ``` -->