---
title: "Kaggle: Model version 1"
always_allow_html: true
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r package, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(zoo)
library(kableExtra)
library(readxl)
library(xts)
```

```{r}
#Importing time series data (relative humidity and temperature optional but importing)

#load data: this is daily data, with 24 different sensors
load_data <- read_excel(
  path = "./Data/load.xlsx",
  sheet = 1
)

#relative humidity: hourly data, with 28 different sensors
relative_humidity_data <- read_excel(
  path = "./Data/relative_humidity.xlsx",
  sheet = 1
)

#temperature data: hourly data, with 28 different sensors
temperature_data <- read_excel(
  path = "./Data/temperature.xlsx",
  sheet = 1
)

#Preparing load data - create date object
load_processed <- load_data %>%
  mutate(date = ymd(date)) %>%
  rowwise() %>%
  mutate(avg_load = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup()

#Turning hourly data into daily data and averaging across row
relative_humidity_processed <- relative_humidity_data %>%
  mutate(date = ymd(date)) %>%
  rowwise() %>%
  mutate(avg_humidity = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  group_by(date) %>%
  summarise(daily_avg_humidity = mean(avg_humidity, na.rm = TRUE)) %>%
  ungroup()

#Turning hourly data into daily data and averaging across row
temperature_processed <- temperature_data %>%
  mutate(date = ymd(date)) %>%
  rowwise() %>%
  mutate(avg_temp = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  group_by(date) %>%
  summarise(daily_avg_temp = mean(avg_temp, na.rm = TRUE)) %>%
  ungroup()

ts_humidity <- ts(relative_humidity_processed$daily_avg_humidity,
                           start=c(2005,1),
                           frequency=365) 

ts_temperature <- ts(temperature_processed$daily_avg_temp,
                           start=c(2005,1),
                           frequency=365) 

relative_humidity_vector <- relative_humidity_processed %>%
  rowwise() %>%
  mutate(avg_rh = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>%
  pull(avg_rh)

temperature_vector <- temperature_processed %>%
  rowwise() %>%
  mutate(avg_temp = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>%
  pull(avg_temp)

summary(load_processed)

#There are 6 NAs in h2

#Turn load into time series data
n_h <- ncol(load_processed)-2
ts_load_daily <- msts(load_processed$avg_load, seasonal.periods = c(7,365.25), start = c(2005, 1,1))
```

#Creating training and testing subsets
```{r message=FALSE, warning=FALSE}
n_for <- 365

# Get number of time steps
n_obs <- length(ts_load_daily)

#Get cutoff point for train/test
cut_point <- n_obs - n_for

# TRAIN SET 
ts_load_train <- window(ts_load_daily, end = time(ts_load_daily)[cut_point])

# TEST SET
ts_load_test <- window(ts_load_daily, start = time(ts_load_daily)[cut_point + 1])
```

#Plotting Data
```{r message=FALSE, warning=FALSE}
autoplot(train_ts, series = "Train") +
      autolayer(test_ts, series = "Test") +
      ggtitle(paste("Train vs Test for", colnames(ts_load_daily)[i])) +
      ylab("Load") +
      xlab("Time")
```

# Model 1: STL + ETS
```{r}
#Fit and forecast STL + ETS model to data
ETS_fit <-  stlf(ts_load_train,h=365)

#Plot foresting results
autoplot(ETS_fit) + ylab("Load")

#Plot model + observed data
autoplot(ts_load_train) +
  autolayer(ETS_fit, series="STL + ETS",PI=FALSE) +
  ylab("Load")

#alternative coding if you don't need to save the objects
#ts_act_power_daily_train %>% stlf(h=365) %>% autoplot() 
#very easy to read/run/implement 
#but you don't have anything stored on your environment

```
# Model 2: ARIMA + FOURIER terms
```{r ARIMA, echo=TRUE, message=FALSE, warning=FALSE}
#Fit arima model with fourier terms as exogenous regressors
# seasonal = FALSE is the same as P=D=Q=0
# play with K by changing it to K=c(2,2), K=c(2,4), K=c(2,6), etc. The higher teh K the longer it will take to converge, because R will try more models.

ARIMA_Four_fit <- auto.arima(ts_load_train, 
                             seasonal=FALSE, 
                             lambda=0,
                             xreg=fourier(ts_load_train, 
                                          K=c(2,4))
                             )

#Forecast with ARIMA fit
#also need to specify h for fourier terms
ARIMA_Four_for <- forecast(ARIMA_Four_fit,
                           xreg=fourier(ts_load_train,
                                        K=c(2,4),
                                        h=365),
                           h=365
                           ) 

#Plot foresting results
autoplot(ARIMA_Four_for) + ylab("Load")

#Plot model + observed data
autoplot(ts_load_train) +
  autolayer(ARIMA_Four_for, series="ARIMA_FOURIER",PI=FALSE) +
  ylab("Load")

```

# Model 3: TBATS
```{r TBATS, echo=TRUE, message=FALSE, warning=FALSE}
# TBATS can take time to fit
TBATS_fit <- tbats(ts_load_train)

TBATS_for <- forecast(TBATS_fit, h=365)

#Plot foresting results
autoplot(TBATS_for) +
  ylab("Load") 

#Plot model + observed data
autoplot(ts_load_train) +
  autolayer(TBATS_for, series="TBATS",PI=FALSE)+
  ylab("Load") 
```

#Check to see how well the forecasted data aligns with the test data, looking for a higher accuracy here
```{r}
accuracy(ETS_fit, ts_load_test)
accuracy(ARIMA_Four_for, ts_load_test)
accuracy(TBATS_for, ts_load_test)
```
#With the best model, fit on the full dataset and forecast the next 59 days
```{r}
TBATS_fit <- tbats(ts_load_daily, seasonal.periods = c(7, 365.25))
TBATS_for <- forecast(TBATS_fit, h=365)

autoplot(TBATS_for) +
  ylab("Load") 

tbats_forecast_df <- data.frame(
  date = time(TBATS_for$mean),
  load = as.numeric(TBATS_for$mean)
)
write.csv(tbats_forecast_df, file = "submission_tbats.csv", row.names = FALSE)
```

#Submitting with ARIMA (in order to add some exogenous variables)
```{r}
fourier_terms <- fourier(ts_load_daily, K = c(2, 4))
# Combine Fourier terms and humidity into one xreg matrix
xreg_combined <- cbind(fourier_terms, RH = relative_humidity_vector, temp = temperature_vector)

ARIMA_Four_fit <- auto.arima(ts_load_daily,
                             seasonal = FALSE,
                             lambda = 0,
                             xreg = xreg_combined
                             )

#Forecast with ARIMA fit
#Need to forecast relative humidity and temperature

# Auto seasonal ARIMA for humidity
fit_humidity <- auto.arima(ts_humidity, seasonal = TRUE)

# Auto seasonal ARIMA for temperature
fit_temperature <- auto.arima(ts_temperature, seasonal = TRUE)

# Forecasting (e.g., 12 months ahead)
forecast_humidity <- forecast(fit_humidity, h = 59)
forecast_temperature <- forecast(fit_temperature, h = 59)

#also need to specify h for fourier terms
future_fourier <- fourier(ts_load_daily, K = c(2, 4), h = 59)
xreg_future <- cbind(future_fourier, RH = future_rh, temp = future_temp)

ARIMA_Four_for <- forecast(
  ARIMA_Four_fit,
  xreg = xreg_future,
  h = 59
)

autoplot(ARIMA_Four_for) + ylab("Load")

ARIMA_Four_forecast_df <- data.frame(
  date = time(ARIMA_Four_for$mean),
  load = as.numeric(ARIMA_Four_for$mean)
)

write.csv(ARIMA_Four_forecast_df, file = "submission_ARIMA.csv", row.names = FALSE)
```

#ARIMA without exogenous variable
```{r}
ARIMA_Four_fit <- auto.arima(ts_load_daily, 
                             seasonal=FALSE, 
                             lambda=0,
                             xreg=fourier(ts_load_daily, 
                                          K=c(2,4))
                             )

#Forecast with ARIMA fit
ARIMA_Four_for <- forecast(ARIMA_Four_fit,
                           xreg=fourier(ts_load_daily,
                                        K=c(2,4),
                                        h=59),
                           h=59
                           ) 

autoplot(ARIMA_Four_for) + ylab("Load")

ARIMA_Four_forecast_df <- data.frame(
  date = time(ARIMA_Four_for$mean),
  load = as.numeric(ARIMA_Four_for$mean)
)

write.csv(ARIMA_Four_forecast_df, file = "submission_ARIMA.csv", row.names = FALSE)
```

#STL + ETS forecast (did not do great!)
```{r}
#Fit and forecast STL + ETS model to data
ETS_fit <-  stlf(ts_load_daily,h=59)

#Plot foresting results
autoplot(ETS_fit) + ylab("Load")

#Plot model + observed data
autoplot(ts_load_daily) +
  autolayer(ETS_fit, series="STL + ETS",PI=FALSE) +
  ylab("Load")

forecast_df <- as_tibble(ETS_fit)
write.csv(forecast_df, file = "submission_stlf.csv", row.names = FALSE)
```
