---
title: "NN Networks"
always_allow_html: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r package, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(zoo)
library(kableExtra)
library(readxl)
library(xts)
```

#Data Import and Primary Cleaning
```{r}
#Importing time series data (relative humidity and temperature optional but importing)

#load data: this is daily data, with 24 different sensors
load_data <- read_excel(
  path = "./Data/load.xlsx",
  sheet = 1
)

#There are 6 NA  andd 17 zeros in data. They would have low impact in the overall
sum(is.na(load_data))
sum(load_data == 0, na.rm = TRUE)

#relative humidity: hourly data, with 28 different sensors
#I would estimate a mean 
relative_humidity_data <- read_excel(
  path = "./Data/relative_humidity.xlsx",
  sheet = 1
)

#There are 0 NA and 30 zeros. They would have low impact
sum(is.na(relative_humidity_data))
sum(relative_humidity_data == 0, na.rm = TRUE)

#temperature data: hourly data, with 28 different sensors
temperature_data <- read_excel(
  path = "./Data/temperature.xlsx",
  sheet = 1
)

#There are 0 NA and one zero. They would have low impact
sum(is.na(temperature_data))
sum(temperature_data == 0, na.rm = TRUE)

```

```{r}
#Processing load data - wrangling to allow the estimation per year, month, day, hour
load_processed <- load_data %>%
  pivot_longer(
    cols = starts_with("h"), 
    names_to = "hour", 
    values_to = "electricity_demand"
  ) %>%
  mutate(
    hour = as.integer(sub("h", "", hour)),  # Extract numeric part from "h1", "h2", etc.
    date = ymd(date),  
    year = year(date), 
    month = month(date), 
    day = day(date)
  ) %>% 
  select(date, year, month, day, hour, electricity_demand)

# Processing Humidity Data: wrangling to allow the estimation per year, month, day, hour
humidity_processed <- relative_humidity_data %>%
  pivot_longer(
    cols = starts_with("rh"), 
    names_to = "hour", 
    values_to = "relative_humidity"
  ) %>%
  mutate(
    hour = as.integer(gsub("[^0-9]", "", hour)),  # Extract only numeric characters (e.g., "rh_ws15" becomes 15)
    date = ymd(date),  
    year = year(date), 
    month = month(date), 
    day = day(date)
  ) %>% 
  select(date, year, month, day, hour, relative_humidity)

# Processing Temperature Data: wrangling to allow the estimation per year, month, day, hour
temp_processed <- temperature_data %>%
  pivot_longer(
    cols = starts_with("t"), 
    names_to = "hour", 
    values_to = "temperature"  # Corrected: use "temperature" instead of "relative_humidity"
  ) %>%
  mutate(
    hour = as.integer(gsub("[^0-9]", "", hour)),  # Extract numeric part
    date = ymd(date),  
    year = year(date), 
    month = month(date), 
    day = day(date)
  ) %>% 
  select(date, year, month, day, hour, temperature)
```
## Converting to daily by taking averages

```{r}
# Converting each processed dataset to daily values
daily_load <- load_processed %>%
  group_by(date) %>%
  summarise(daily_avg_load = mean(electricity_demand, na.rm = TRUE)) %>%
  ungroup()

daily_humidity <- humidity_processed %>%
  group_by(date) %>%
  summarise(daily_avg_humidity = mean(relative_humidity, na.rm = TRUE)) %>%
  ungroup()

daily_temp <- temp_processed %>%
  group_by(date) %>%
  summarise(daily_avg_temp = mean(temperature, na.rm = TRUE)) %>%
  ungroup()

```
## Merging the daily datasets to a full daily dataset 

```{r}
# Merging the daily datasets into one full dataset by date
full_daily <- daily_load %>%
  inner_join(daily_temp, by = "date") %>%
  inner_join(daily_humidity, by = "date") %>%
  arrange(date)

# Preview the merged full dataset
head(full_daily)
```
##Converting to time series object

```{r message=FALSE, warning=FALSE}
# Create the daily time series object from the merged full_daily dataset
# I use msts() to capture both weekly (7-day) and annual (365.25-day) seasonal patterns.
ts_electricity_daily <- msts(full_daily$daily_avg_load, 
                              seasonal.periods = c(7, 365.25), 
                              start = decimal_date(as.Date("2005-01-01")))

# I plot the full daily series to verify it looks as expected.
autoplot(ts_electricity_daily) + ggtitle("Electricity Demand: Daily")

# Now, I split the time series into training and test sets.
# The training set is from January 1, 2005, to December 31, 2009.
# The test set (for validation) covers January 1 to February 28, 2010.
# I use the window() function to create these subsets.

ts_daily_train <- window(ts_electricity_daily, end = c(2009, 365))
ts_daily_test  <- window(ts_electricity_daily, start = c(2010, 1), end = c(2010, 59))

# I visualize the training and test sets to confirm the split.
autoplot(ts_daily_train) + ggtitle("Training Set: Daily Demand (2005-2009)")
autoplot(ts_daily_test) + ggtitle("Test Set: Daily Demand (Jan-Feb 2010)")
```

#Neural Network Models

##p = 2, P = 2, K = c(2,12)
```{r ETS, echo=TRUE, message=FALSE, warning=FALSE}
horizon <- length(ts_daily_test)

NN_fit <- nnetar(ts_daily_train,
                  p=2,
                  P=2,
                  xreg=fourier(ts_daily_train, K=c(2,12)))

NN_for <- forecast(NN_fit, h=horizon,xreg=fourier(ts_daily_train, 
                                           K=c(2,12),h = horizon))

#Plot model + observed data
autoplot(ts_daily_train) +
   autolayer(NN_for, series="Neural Network",PI=FALSE) +
   ylab("Load") 

accuracy_NN_for <- accuracy(NN_for, ts_daily_test)
accuracy_NN_for
```

               ME         RMSE       MAE        MPE      MAPE      MASE     ACF1 Theil's U
Training set  -0.8713464  154.466  111.3658 -0.4865963  3.371852 0.1447422 -0.1308867        NA
Test set     641.3090394 1942.737 1600.4780  5.1390523 31.937919 2.0801424  0.8493126  2.370443

##p = 6, P = 3, K = c(2,12)
Increasing the number of past lags and seasonal lags to increase the amount of past information.
```{r ETS, echo=TRUE, message=FALSE, warning=FALSE}
horizon <- length(ts_daily_test)

NN_fit <- nnetar(ts_daily_train,
                  p=6,
                  P=3,
                  xreg=fourier(ts_daily_train, K=c(2,12)))

NN_for <- forecast(NN_fit, h=horizon,xreg=fourier(ts_daily_train, 
                                           K=c(2,12),h = horizon))

#Plot model + observed data
autoplot(ts_daily_train) +
   autolayer(NN_for, series="Neural Network",PI=FALSE) +
   ylab("Load") 

accuracy_NN_for <- accuracy(NN_for, ts_daily_test)
accuracy_NN_for
```
The error on the test set seems to have gone up, so probably will not be using this model.

                      ME       RMSE        MAE         MPE     MAPE       MASE       ACF1 Theil's U
Training set   0.1183327   31.69454   22.30719 -0.06463893  0.71525 0.02899267 -0.2279991        NA
Test set     928.1271916 2137.16583 1837.29319 10.94731840 36.38336 2.38793125  0.8417433  2.593319

##p = 2, P = 2, K = c(3,15)
For the first seasonality, using 5 harmonics. For the second seasonality, using 15 harmonics.

```{r ETS, echo=TRUE, message=FALSE, warning=FALSE}
NN_fit <- nnetar(ts_daily_train,
                  p=2,
                  P=2,
                  xreg=fourier(ts_daily_train, K=c(3, 15)))

NN_for <- forecast(NN_fit, h=horizon,xreg=fourier(ts_daily_train, 
                                           K=c(3,15),h = horizon))

#Plot model + observed data
autoplot(ts_daily_train) +
   autolayer(NN_for, series="Neural Network",PI=FALSE) +
   ylab("Load") 

accuracy_NN_for <- accuracy(NN_for, ts_daily_test)
accuracy_NN_for
```
The error on the test set went up here, so probably will not stick with this.

                      ME      RMSE        MAE        MPE     MAPE       MASE       ACF1 Theil's U
Training set  -0.7055144   73.2614   51.20423 -0.1823739  1.59530 0.06655017 -0.1012514        NA
Test set     789.7083269 2003.7694 1766.25663  8.1477621 35.52517 2.29560498  0.8344838  2.488114

##p = 2, P = 2, K = c(2,13)
For the first seasonality, using 5 harmonics. For the second seasonality, using 15 harmonics.

```{r ETS, echo=TRUE, message=FALSE, warning=FALSE}
NN_fit <- nnetar(ts_daily_train,
                  p=2,
                  P=2,
                  xreg=fourier(ts_daily_train, K=c(2, 13)))

NN_for <- forecast(NN_fit, h=horizon,xreg=fourier(ts_daily_train, 
                                           K=c(2,13),h = horizon))

#Plot model + observed data
autoplot(ts_daily_train) +
   autolayer(NN_for, series="Neural Network",PI=FALSE) +
   ylab("Load") 

accuracy_NN_for <- accuracy(NN_for, ts_daily_test)
accuracy_NN_for
```

                      ME      RMSE        MAE        MPE      MAPE      MASE       ACF1 Theil's U
Training set  -0.4984742  118.6062   85.68213 -0.3390185  2.595967 0.1113611 -0.1028054        NA
Test set     730.8305853 2018.3051 1715.24692  6.7696245 34.261757 2.2293076  0.8349024  2.470025

##p = 2, P = 2, K = c(2,8)
Trying to improvegeneralization

```{r ETS, echo=TRUE, message=FALSE, warning=FALSE}
NN_fit <- nnetar(ts_daily_train,
                  p=2,
                  P=2,
                  xreg=fourier(ts_daily_train, K=c(2, 8)))

NN_for <- forecast(NN_fit, h=horizon,xreg=fourier(ts_daily_train, 
                                           K=c(2,8),h = horizon))

#Plot model + observed data
autoplot(ts_daily_train) +
   autolayer(NN_for, series="Neural Network",PI=FALSE) +
   ylab("Load") 

accuracy_NN_for <- accuracy(NN_for, ts_daily_test)
accuracy_NN_for
```

Some better error metrics on test set, but not for all, still worth trying to run it.

## Final forecast for best neural networks of interest
```{r}
# submission forecast horizon: 59 days (from Jan 1, 2011 to Feb 28, 2011)
horizon_final <- 59

NN_fit_22_final <- nnetar(ts_electricity_daily,
                  p=2,
                  P=2,
                  xreg=fourier(ts_electricity_daily, K=c(2,12)))
NN_for_22_final <- forecast(NN_fit_22_final, h=horizon_final,xreg=fourier(ts_daily_train, 
                                           K=c(2,12),h = horizon_final))

NN_fit_28_final <- nnetar(ts_electricity_daily,
                  p=2,
                  P=2,
                  xreg=fourier(ts_electricity_daily, K=c(2,8)))
NN_for_28_final <- forecast(NN_fit_28_final, h=horizon_final,xreg=fourier(ts_daily_train, 
                                           K=c(2,8),h = horizon_final))
```

## Saving the forecasts
```{r}

template <- read_excel("./Data/load.xlsx")

# NN forecast according to template
final_forecast_NN_df <- data.frame(
   date = time(NN_for_22_final$mean),
   load = as.numeric(NN_for_22_final$mean))

write.csv(final_forecast_NN_df, file = "submission_forecast_NN.csv", row.names = FALSE)

# NN forecast according to template
final_forecast_NN_28_df <- data.frame(
   date = time(NN_for_28_final$mean),
   load = as.numeric(NN_for_28_final$mean))

write.csv(final_forecast_NN_28_df, file = "submission_forecast_NN_28.csv", row.names = FALSE)

```
